---
title: "IanBachli - Coding Exercise, Updated for Tidyverse by Megan Beaudry"
author: Ian Bachli & Megan Beaudry
Date: 08/27/2019
Update 2: 9-16-19
output:
  html_document:
    df_print: paged
  word_document: default
---

```{r}
# Loaded the dslabs library followed by the gapminder data set. 
library(dslabs)

#Gapminder is a collection of data containing health and income outcomes for 184 countries ranging from the years 1960 to 2016.
data(gapminder)
```

#For the tidyverse exercise we need first install packages as we have not had them previously installed
#load the libraries for tidyverse and skimr
```{r}
library(tidyverse)
library(skimr)
library(dplyr)
```

#Now we want to take glimpse of the data. We can do this using the glimpse function to look at the gapminder dataset
```{r}
glimpse(gapminder)
```
#the glimpse function tells us we have 10,545 observations for 9 variables.

#skim function from dyplry to look at the gapminder data
```{r}
skim(gapminder)
```
#the purpose of skim is to generate a summary of the data. It also provides with the number of observations and variables. However, it gives us further information on the missing values

```{r}
#To view the the information in the gapminder data set, use the following R functions: help, str, summary, and class. Below is a summary of each.

help(gapminder)
#The help function provides help pages about the selected package within the help tab.

str(gapminder)
#Structure, or the str function, displays the basic structural information for an object in the console. Using the str function we can see that gapminder consists of nine variables: country, infant_mortality, life_expectancy, fertility, population, gdp, continent, and region. 

summary(gapminder)
#The summary function will display statistical summaries of the selected data in the console for each of the variables within the data set. The type of data shown will depend on the class (i.e. numeric, integer, character, etc.) of the individual variables within a data set. 

class(gapminder)
#The class function will display the type of object that is represented by a given name. For example, gapminder is of the data.frame class. 
```

```{r}
#Assign only the African countries to a new object titled africadata.

africadata <- gapminder$continent == "Africa"
#To assign a new object use the <- to the right of the chosen object name. All African countries can be accessed in gapminder through the continent variable. The accessor $ is used to access specific variables within the data frame. To determine if a country is within Africa, africadata is assigned as a logical vector using == to determine if the continent variable is listed as "Africa" or other. 

str(africadata)
#The logical vector africadata will output information as TRUE or FALSE depending on the continent variable of each country. In this example, 10545 different data points are measured. Countries in Africa receive the output TRUE whereas all other continents receive a FALSE. 

summary(africadata)
#The summary shows results of the logical vector. Of the total 10545 measured countries, 2907 of them are located in Africa. These values can be verified by reviewing the previous summary(gapminder) output in the console and subtracting the African data points from the total measurements under the continent variable. 
```

```{r}
#Create an object containing infant_mortality and life_expectancy for africadata.

#Create and object for infant_mortality for African countries by accessing gapminder with $ and applying the logical vector africadata.
infant_mortality_africa <- gapminder$infant_mortality[africadata]

#Repeat the same process to create an object fot life_expectancy for African countries.
life_expectancy_africa <- gapminder$life_expectancy[africadata]

#Combine the two new objects into a data frame to create a single object containing both variables. 
infant_vs_life_africa <- data.frame(infantmortality = infant_mortality_africa, lifeexpectancy = life_expectancy_africa)

#The result is a data frame with 2907 observations and two variables: infantmortality and lifeexpectancy.
str(infant_vs_life_africa)

#Summary gives a quick glance of min, max, median, and mean for both variables. Note there are 226 NA records indicated for infantmortality.
summary(infant_vs_life_africa)
```
#Extract only the african countries from the gapminder dataset using dplyr functions
#convert the data to a tibble to view
```{r}
africancountries <- filter(gapminder, continent == "Africa")
africatibble <- tibble(africancountries)
africatibble
```
#Now we want to select for the following variables: infant_mortality, life_expectancy, population, and country.
```{r}
africa_plot_data <- select(africancountries, life_expectancy, infant_mortality, population, country)
africa_plot_data
```
#Result in 2,907 observations and only the four variables we selected for
#Another option for selecting this would be to place a "-" sympbol in front of each of the variables we did NOT WANT, however in this case it was easier to list the ones we did want

```{r}
#Create an object containing population and life_expectancy for africadata.

#Repeat the process above for the population variable.
population_africa <- gapminder$population[africadata]

#Combine the population and life_expectancy data into a data frame.
pop_vs_life_africa <- data.frame(population = population_africa, lifeexpectancy = life_expectancy_africa)

#The result is a data frame with 2907 observations and two variables: population and lifeexpectancy.
str(pop_vs_life_africa)

#Standard statistical values are given in the summary of the object. Note 51 NA values were recorded for the population variable.
summary(pop_vs_life_africa)
```

```{r}
#Plot of infant mortality verses life expectancy for all African countries.
plot(infant_vs_life_africa)
#The plot illistrates a distinct negative correlation between the two variables. As expected, countries with increased infant mortality show decreased life expectancy.
```

```{r}
#Plot of population verses life expectancy for all African countries. 
#Convert previous population object into a new object at log scale.
logten_pop_africa <- log10(population_africa)
#Input the new log population data into a data frame with life expectancy. 
logtenpop_vs_life_africa <- data.frame( log10population = logten_pop_africa, lifexcpectancy = life_expectancy_africa)
plot(logtenpop_vs_life_africa)
#The plot shows a distinct scattering of "streaks" of data. The individual streaks typically move in a positive trend illistrating a positve correlation between increased population size and increased life expectancy. The "streaks" seen in the data likely represent different contries within the continent of Africa measured over time. 
```
##Make 2 plots using ggplot for life expectancy as a function of infant mortality and population. Also, give each country a separate color in the plot.
#In order to figure out which plotting functions to use i went to this page: https://ggplot2.tidyverse.org/reference/translate_qplot_ggplot.html. It provides useful comparisons
##They explained that qplot is best for interactive use or quicker plots
##ggplot may be more effective for complex figures 

#First plot: Infant mortality vs life expectancy (x vs y)
#synatx - x verse y function, color defines the data point color, and data assigns the object. Labs fucntion gives us the labels, the them function setts the formating 
```{r}
help("ggplot2")
qplot(infant_mortality, life_expectancy, color= country, data = africa_plot_data) + labs(y = "Life Expectancy", x = "Infant Mortality") + theme(legend.key.size = unit(0.2, "cm"), legend.key.width = unit(0.1, "cm"))
```
#This shows the same negative correlation that was seen previously, however it is color coded!
#Warning removal of 226 values is consistent with the number of NA values for infant mortality

#Plot 2 - population vs life expectancy
#Population NEEDS to be set to log scale
```{r}
qplot(population, life_expectancy, color = country, data = africa_plot_data) + labs(y = "Life Expectancy", x = "Population (log10)") + scale_x_log10() + theme(legend.key.size = unit(0.2, "cm"), legend.key.width = unit(0.1, "cm"))
```
#Here we see the same "smears" we saw previously, but again it is color coded. 
#We can improvise from this graph that as life expectancy increases population also increases 
#Warning removal of 51 rows is consistent with the number of NA values for population

#Plot 3
#Write code that uses the gapminder object, pulls out data for africa and year 2000 and creates a ggplot for life_expectancy as a function of infant_mortality, with a linear model fit added.
#First, we need to filter out the variables we want from the gapminder dataset
```{r}
africa2000 <- filter(gapminder, continent == "Africa" & year == 2000)
```
#Now plot infant mortality vs life expectancy with linear fit model
#ggplot function we will define africa2000 as the data and set the axes
#geopointd defines the scatterplot
#stat_smooth applies the linear fit with "lm" being the method used
#col sets the color of the of the regression line
#se applies or removes the standard error
```{r}
ggplot(africa2000, aes(x = infant_mortality, y = life_expectancy, color = country)) + geom_point() + stat_smooth(method = "lm", col = "red", se = FALSE) + labs(y = "Life Expectancy", x = "Infant Mortality") + theme(legend.key.size = unit(0.2, "cm"), legend.key.width = unit(0.1, "cm"))
```
#this plot reflects a distinct negative correlation between life expectancy and infant mortality

```{r}
#Determine which years have missing data for infant mortality.
#Create an index object for the NA values within the infant_mortality_africa object using the is.na function. This produces a logical vector that tells us which entries are NA. 
na_index <- is.na(infant_mortality_africa)

#Apply the na_index to the gapminder year data.
gapminder$year[na_index]

#The result is a list of years where infant mortality data is missing. Repetitive years in the list represent years where infant mortality data is missing from multiple countries within the continent of Africa.

#The sum function checks the total recorded Na values within the na_index to ensure it is consistent the previous summary of infant mortality.
sum(na_index)
```

```{r}
#Create a new object by extracting the data from the year 2000 from africadata. 
#Create an index to select for the year 2000.
year2000 <- gapminder$year == "2000"

#Combine the new index with africadata to select for Africa and the year 2000 combined.
africa2000 <- africadata & year2000

#Apply this index to all variables in the gapminder data set.  
year2000 <- gapminder$year[africa2000] 
region2000 <- gapminder$region[africa2000]
country2000 <- gapminder$country[africa2000]
infantmort2000 <- gapminder$infant_mortality[africa2000]
lifeexpect2000 <- gapminder$life_expectancy[africa2000]
fertility2000 <- gapminder$fertility[africa2000]
gdp2000 <- gapminder$gdp[africa2000]
continent2000 <- gapminder$continent[africa2000]
population2000 <- gapminder$population[africa2000]

#Create a data frame using the new object names to select for African countries and the year 2000. 
africa2000df <- data.frame(year = year2000, country = country2000, population = population2000, region = region2000, infant_mortality = infantmort2000, life_expectancy = lifeexpect2000, fertility = fertility2000, gdp = gdp2000, continent = continent2000)

str(africa2000df)
#The resulting data frame has 51 observations and 9 vairables. 

summary(africa2000df)
#Specific statistical values are given for each of the 9 variables within the object.
```

```{r}
#Recreate the plot for infant mortality verses life expectancy.
plot(infantmort2000, lifeexpect2000)

#The resulting figure shows a negative correlation between infant mortality and life expectancy. 
```

```{r}
#Recreate the plot for population verses life expectancy. 
#Redefine the population data for Africa in 2000 to a log 10 scale. 
logtenpop2000 <- log10(population2000)
#Plot log 10 population verses life expectancy.
plot(logtenpop2000, lifeexpect2000)
#The resulting figure shows no distinct correlation between population and life expectancy. 
```

```{r}
#Create a linear regression model for life expectancy as the outcome (y variable), and infant mortality as the predictor (x variable).Ensure the y variable is the first variable entered to the lm() function.
fit1 <- lm(lifeexpect2000 ~ infantmort2000)
 
summary(fit1)
#Summary prints the fit results to the console and provides a number of  important statistical data points, in particular the p-value. The p-value from this example is 2.826e-08. The very low p-value suggests that infant mortality is strongly related to life expectancy. 
 
```

```{r}
#Create a linear regression model for life expectancy as the outcome (y variable), and population size as the predictor (x variable).
fit2 <- lm(lifeexpect2000 ~ logtenpop2000)
summary(fit2)
#The p-value from this example is 0.4755. The high p-value suggests that poplation size is not related to life expectancy.
```

```{r}
#Fit line to scatterplot for life expectancy verses infant mortality.
#Replot the scatterplot.
plot(infantmort2000, lifeexpect2000)
#Use the abline function to fit a line to the plot according to the new linear model object. 
abline(fit1)
#Here we see a strong negative correlation with the fit line which is consistent with the p-value.
```

```{r}
#Fit line to scatterplot for life expectancy verses population size.
plot(logtenpop2000, lifeexpect2000)
abline(fit2)
#Here we see no distinct correlation with the fit line which is consistent with the p-value. 
```